---
title: "Practice Problems #2"
author: "Alvin D. Jeffery, PhD, RN"
date: "Last Updated: `r Sys.Date()`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, error = FALSE, warning=FALSE, comment="", echo = FALSE)
library(Hmisc); library(data.table); library(dplyr); library(tidyr)
```


# Load and Review Data

This is a dataset that Maddie had shared with us.  (Maddie can provide overview.)

Begin by reading in and exploring the data.  

```{r}
### read/load the data

### explore the data


```

### Possible Approach
```{r eval=FALSE}
# answer key
w <- fread('../data_wrangling/maddie/brain_fu_2015-04-28.csv')
str(w)
```

## Filter by Follow-Up Time

Let's say we want to only look at certain follow-up times.  

```{r}
### identify all possible values of 'fu.period'


### subset 'fu.period' to '3 Month'

### create a new CSV file with the filtered data 

```

### Possible Approach

```{r eval=FALSE}
# answer key
unique(w$fu.period) # identify possible values
three_months <- filter(w, fu.period == '3 Month') # subset to specified time
write.csv(three_months, # create new CSV file
          file = '../data_wrangling/maddie/brain_fu_three_months.csv') 
```

# Working with paste()

Notice how long these file paths (directories) are.  It can be a burden to have to continue typing them again and again.  Additionally, it makes the code a little messy.  Let's work on storing the directory in a variable and then calling it in future functions.  

```{r eval=FALSE}
?paste # check out the help file first

# store path/directory in a variable
directory <- c('../data_wrangling/maddie/')
directory # see how it looks by itself

paste(directory, 'brain_fu_three_months.csv', sep='') # use paste w/filename & extension

```

This could also help us quickly filter & save files for multiple follow-up times.  Let's pretend we want to filter our data to each unique time point, and we want a separate file for each.  Performing this task could be as simple as embedding 2 lines of code within a for-loop.  

```{r eval=FALSE}
### enter your own code here (comments provided for each line)

for (i in values) {  # begin for-loop [hint: consider using unique() ]
  # filter data to specified fu.period timeframe 

  # write filtered data to file using paste() to create unique file names 

}  # end for-loop

```

### Possible Approach
```{r eval=FALSE}

for (i in unique(w$fu.period)) {
  data <- filter(w, fu.period == i) # subset/filter to each time period
  write.csv(data, # write filtered data to new file
            file = paste(directory, 'brain_fu_', i, '.csv', sep=''))
  
  # if saving as .rdata file
  save(data, file = paste(directory, 'brain_fu_', i, '.rdata', sep=''))
}

```

# Filter by First Study Date

In the other file, we want to subset each patient to only their first study day (i.e., find baseline information).  

```{r eval=FALSE}
# I am providing my own code because working with dates is SOOOO challenging
# you really want to be connected to the internet to troubleshoot your error messages :-)  
v <- fread(paste(directory, 'brain_assess_2015-04-28.csv', sep=''),
           data.table = FALSE)
str(v)
# notice how dates are listed as character variables - we want to change this

v <- cleanup.import(v, 
                    # datetimevars = 'asmt.posix', # can't get to work
                    # dateformat = '%m/%d/%Y %H:%M', 
                    datevars = 'asmt.date',
                    dateformat = '%m/%d/%Y')
str(v) # did it work?

first_day <- v %>% # prepare to store in new object while working on code
  arrange(asmt.date) %>% # place all dates in ascending order (optional)
  group_by(id) %>% # ensures the next step is done unique for each patient id
  mutate(first.date = min(asmt.date)) %>% # add new column with earliest date
  filter(asmt.date == first.date) %>% # remove all dates that aren't the first
  select(-first.date) # remove the intermediary variable (optional)


```

# Removing Duplicates

One last thing I had to do is sort out duplicate study ID’s and also how to narrow down the list to only a specific set of patients. For example, the BRAIN data has 821 patients, but this idea density project only has ~240. How to get that 821 to only the ~240.
 
```{r eval=FALSE}
str(w) # brain fu 
str(v) # brain assess 

library(foreign)
u <- read.spss(file = paste(directory, 'ideadenseids.sav', sep = ''), 
               to.data.frame = TRUE)

```



