---
title: "Practice Problems #3"
author: "Alvin D. Jeffery, PhD, RN"
date: "Last Updated: `r Sys.Date()`"
output: 
  pdf_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, error = FALSE, warning=FALSE, comment="", echo = TRUE)
library(Hmisc); library(data.table); library(dplyr); library(tidyr)
```

```{r make_bad_data, echo=FALSE, eval=FALSE}
# Alvin's code for making a 'bad' data set that we can work on preparing
getHdata(support) # load original file
# add patient id for separating & re-joining later
set.seed(123)
support$id <- base::sample(1e4:1e5, size = nrow(support))
# make separate vital signs, lab values, & demographics files
vitals <- select(support, id, temp, hrt, resp, meanbp)
labs <- select(support, id, wblc, alb, bili, crea, sod, ph, glucose, bun)
demographics <- select(support, id, age, sex, edu, income, dzgroup)

# make messy data
vitals$hrt <- as.character(vitals$hrt) # coerce numeric to character

Save(vitals); Save(labs); Save(demographics)
```

The purpose of this problem set is to practice multiple aspects of wrangling and analyzing data.  You will open up 3 separate datasets coming from the SUPPORT dataset, explore & manipulate them, and then do some analytical work.  

# Load and Review Data

Begin by reading in and exploring the data.  

```{r}
### read/load the data


### explore the data


```


```{r echo=FALSE, eval=FALSE}
# answer key
Load(vitals); Load(labs); Load(demographics)
```

# Descriptive Statistics

Explore the following items:  
1. Create a table of how many patients have each disease.  
2. What is the mean heart rate?  
3. Convert temperature from Celsius to Farenheit.  
4. Set non-physiological respiratory rates to missing.  

```{r}
### disease table


### troubleshoot discovering the mean heart rate


### create new variable tempF for temperature in Farenheit


### if respiratory rate < 8, set to NA


```


```{r echo=FALSE, eval=FALSE}
# answer key
table(demographics$dzgroup)

vitals$hrt <- as.numeric(vitals$hrt)
mean(vitals$hrt, na.rm = T)

vitals$tempF <- vitals$temp * 9/5 + 32

vitals$resp <- ifelse(vitals$resp < 8, 
                      NA, 
                      vitals$resp)
```

\newpage
# Combine Datasets

Combine all 3 datasets into 1 big dataframe for conducting analyses.  There should be 1 row per patient.  Then, sort the large dataframe by patient's ages.  

*Challenge:* Re-write all this code as one chain.  

```{r}
### combine demographics and vitals dataframes into an "mydata" object


### add the labs dataframe to the "mydata" object


### sort by age from oldest to youngest


### re-write above lines into 1 chain


```

```{r echo=FALSE, eval=FALSE}
# answer key
mydata <- left_join(demographics, vitals, by='id')
mydata <- left_join(mydata, labs, by='id')
mydata <- arrange(mydata, desc(age))

mydata <- # as a chain
  left_join(demographics, vitals, by = 'id') %>% 
  left_join(labs, by = 'id') %>% 
  arrange(desc(age))
```

# Sex Differences in Heart Rate

Let's say you want to explore sex differences in the heart rate values.  Perform the following procedures:  
1. Create a boxplot figure of the heart rates, grouped by male vs. female  
2. Perform a t-test for inferential analysis  
3. Build a linear model for further inferential analysis  

```{r}
### boxplot (don't forget to add a title & consider rotating your y-axis labels)


### conduct a t-test 


### build a linear model 



```

```{r echo=FALSE, eval=FALSE}
# answer key
boxplot(hrt ~ sex, data = mydata, 
        # plot options
        main = "Heart Rate Differences between Sexes", 
        las = 1)

t.test(hrt ~ sex, data = mydata)

fit <- lm(hrt ~ sex, data = mydata)
summary(fit)
```

# Function for Boxplots

Your co-investigator really loved your beautiful boxplots.  You want to create a function that makes it easier to create these.  Write a function that takes 3 arguments (data, dependent variable, & grouping/independent variable) and produces a boxplot where the title is automatically populated according to the variable names.  

*Hint:* You'll need to use a paste() function within a formula() function in order to pass it appropriately to the boxplot function().  

```{r}
### create function


### attempt some examples


```

```{r echo=FALSE, eval=FALSE}
# answer key
make_boxplots <- function(data = mydata, dv, grouped_by) {
  title_helper <- ifelse(grouped_by == "dzgroup", "Disease Group", 'Sex')
  boxplot(formula(paste(dv, "~", grouped_by)), 
          data = data, 
          main = paste(dv, "differences between", title_helper), 
          las = 1)
}

make_boxplots(dv = 'hrt', 
              grouped_by = 'sex')

# other examples
make_boxplots(dv = 'ph', grouped_by = 'sex')
make_boxplots(dv = 'ph', grouped_by = 'dzgroup')
make_boxplots(dv = 'age', grouped_by = 'dzgroup')
```

